---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# emuhelpeR: Convenience functions for working with emuR

<!-- badges: start -->
<!-- badges: end -->

This package collects convenience functions for working with `emuR` and 
EMU-SDMS. For more information on those tools, see 
[this useful manual](https://ips-lmu.github.io/The-EMU-SDMS-Manual/index.html).

For now, functions are available to bulk extract and pre-process
[SSFF tracks](https://ips-lmu.github.io/The-EMU-SDMS-Manual/chap-sigDataExtr.html#extracting-pre-defined-tracks),
with specific functions available for processing fundamental frequency,
formants, and measures which depend directly on these.
This can all be done in a single step with the function `import_ssfftracks()`,
but the different steps can also be carried out independently.
Some of the independent functions may well be useful for raw data that isn't
generated in `emuR`.
These functions are adapted from the data processing used in 
[Kirby et al. forthc.](https://doi.org/10.17605/OSF.IO/WV6QZ).

## At a glance

The function `import_ssfftracks()` assumes that you have raw data stored
in an EMU database which has already been loaded into R, and that you have
generated a segment list with relevant portions of the data using 
[the querying system](https://ips-lmu.github.io/The-EMU-SDMS-Manual/chap-querysys.html) 
in `emuR`. Let's load some example data from
[Kirby et al. forthc.](https://doi.org/10.17605/OSF.IO/WV6QZ)
into R.

```{r}
datapath <- system.file('extdata/db', package='emuhelpeR')
raw <- emuR::load_emuDB(datapath)
```

This data can be inspected in EMU-SDMS by typing `emuR::serve(raw)` in the R
console. Let's have a look at a segment list I prepared:

```{r}
library(emuhelpeR)
seg_list
dplyr::glimpse(seg_list)
```

There are a bunch of functional measures available for this database, as the
following prompt will tell us:

```{r}
emuR::list_ssffTrackDefinitions(raw)
```

Using `import_ssfftracks()` we can bulk extract all these measures from the
segments in `seg_list` into a single data frame. I set `proc=FALSE`, 
because right now we just want to extract the raw measures. `verbose` is set to
`FALSE` to avoid printing progress bars that look ugly on GitHub.

```{r}
x <- import_ssfftracks(db_handle=raw, seg_list=seg_list, proc=FALSE, verbose=FALSE)
x
dplyr::glimpse(x)
```

Neat! But if we skip `proc=FALSE` and set some more parameters, we can also do
a bunch of preprocessing in the same step, such as by-speaker normalization and
automated removal of outliers that fall outside three standard deviations from
the mean within the same group. When the function is called, it will also print
a message telling us how many outliers were removed from each track.

* `f0col='praatF0'` specifies that F0 values are stored in the SSFF track
`praatF0`. In this track, values of 0 should be recoded as `NA`, and outliers
should be automatically removed after.
* `f0dep='H1H2c'` specifies that the track `H1H2c` (the difference between the
first two harmonics) is directly dependent on the F0 measurements, so for each
F0 measure coded as `NA`, the corresponding `H1H2c` should also be coded as `NA`.
* `fncol=c('praatF1', 'praatF2', 'praatF3')` specifies that the available formant
measures F1-F3 are stored in the SSFF tracks `praatF1`, `praatF2`, and `praatF3`.
Outliers are automatically removed.
* `fndep=list(c('H1A1c', 'F1'), c('H1A3c', 'F3'))` specifies that, respectively,
`H1A1c` is a spectral measure that directly depends on F1 (and F0), and `H1A3c`
is a spectral measure that directly depends on F3 (and F0). `H1A1c` values will
be coded as `NA` if the corresponding F1 *or* F0 measure is `NA`, etc.
* `speaker='speaker'` specifies that there is a column with speaker information
in the `seg_list` data frame, and that column is labeled `speaker`. This is used
for by-speaker normalization. 
* `group_var=c('speaker', 'vowel')` specifies that the columns `speaker` and 
`vowel` in `seg_list` should be used for determining which tokens should be 
automatically removed; only tokens that are three standard deviations from the
mean within-speaker and within-vowel are removed.
* `timing_rm=list('cl', 250)` specifies that F0 measurements that are more than
250 ms removed from a `cl` label in the data should be removed. 
* `outlier_rm='eggF0'` specifies that, in addition to the automated outlier
procedures that have already been applied, outliers should also be automatically
removed from the SSFF track `eggF0`.

```{r}
y <- import_ssfftracks(db_handle=raw, seg_list=seg_list,
                       f0col='praatF0', f0dep='H1H2c', 
                       fncol=c('praatF1', 'praatF2', 'praatF3'),
                       fndep=list(c('H1A1c', 'F1'), c('H1A3c', 'F3')),
                       speaker='speaker', group_var=c('speaker', 'vowel'),
                       timing_rm=list('cl', 250), outlier_rm='eggF0',
                       verbose=FALSE)
y
dplyr::glimpse(y)
```

Notice that the `praatF0`, `praatF1` columns etc. have been renamed to `F0`,
`F1`. Notice also that for each SSFF track has a corresponding column with
z-score normalized values (e.g. `zF1`) *and* a corresponding column where these
normalized values have been rescaled based on the overall mean and standard
deviation of the data (e.g. `normF1`).

`import_ssfftracks()` is very dependent on `emuR` and EMU-SDMS, but it
incorporates several independent functions which can in principle be used on
raw data generated with other software: `f0_proc()` for processing F0 and
dependencies, `fn_proc()` for processing formants and dependencies,
`outlier_rm` for automated removal of outliers, and `normz` for z-score 
normalizing and rescaling by speaker. The syntax of these functions is similar
to `import_ssfftracks()`.

## Installation

You can install the development version of `emuhelpeR` from GitHub with:

```{r}
# install.packages("devtools")
# devtools::install_github("rpuggaardrode/emuhelpeR")
```

## References

Kirby, James, Marc Brunelle & Pittayawat Pittayaporn (forthc.) Transphonologization of onset voicing: Revisiting Northern and Eastern Kmhmu. (To be published in *Phonetica*.). DOI: [10.17605/OSF.IO/WV6QZ](https://doi.org/10.17605/OSF.IO/WV6QZ).
